---
title: "HW1: Importing Data from Different Sources in R"
author: "Zainab Suleiman"
date: "2026-05-07"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

R can import data from many different sources:

1. Text files (TXT, CSV)
2. Excel files (XLSX)
3. Statistical packages: SPSS (.sav), Stata (.dta), SAS (.sas7bdat)
4. Database management systems (SQLite, PostgreSQL, MySQL)

---

## 1. Importing from Text Files

### 1.1 Read a TAB-separated TXT file using base R

```{r read-txt}
# read.table reads any delimiter-separated file
class_data <- read.table("data/class_data.txt", header = TRUE, sep = "\t")
head(class_data)
str(class_data)
```

### 1.2 Read a CSV file

```{r read-csv}
# read.csv is a shortcut for read.table with sep=","
student_data <- read.csv("data/student_data.csv", header = TRUE)
head(student_data)
str(student_data)
```

---

## 2. Importing from Excel Files

```{r install-readxl, eval=FALSE}
install.packages("readxl")
```

```{r read-excel}
library(readxl)

# datasample.xlsx — student sample data
excel_data <- read_excel("data/datasample.xlsx")
head(excel_data)
str(excel_data)

# OpeningDate.xlsx — opening dates dataset
opening_data <- read_excel("data/OpeningDate.xlsx")
head(opening_data)
str(opening_data)
```

---

## 3. Importing from Statistical Packages (SPSS, Stata, SAS)

```{r install-haven, eval=FALSE}
install.packages("haven")
```

```{r load-haven}
library(haven)
```

### 3.1 Import SPSS (.sav) file

```{r spss}
# datasample.sav — SPSS format
spss_data <- read_sav("data/datasample.sav")
head(spss_data)
str(spss_data)
```

### 3.2 Import Stata (.dta) file

```{r stata}
# loan_datasample.dta — loan data in Stata format (100 loan records)
stata_data <- read_dta("data/loan_datasample.dta")
head(stata_data)
str(stata_data)
```

### 3.3 SAS (.sas7bdat) file

```{r sas, eval=FALSE}
# Requires a SAS-format file — example code shown below
# sas_data <- read_sas("data/your_file.sas7bdat")
# head(sas_data)
```

---

## 4. Importing from Database Management Systems

### 4.1 SQLite (portable, no server needed — school.db is included in this project)

```{r install-dbi, eval=FALSE}
install.packages("DBI")
install.packages("RSQLite")
```

```{r sqlite}
library(DBI)
library(RSQLite)

# Connect to the existing SQLite database
con_sqlite <- dbConnect(RSQLite::SQLite(), "data/school.db")

# List available tables
dbListTables(con_sqlite)

# Import the students table
students_db <- dbGetQuery(con_sqlite, "SELECT * FROM students")
head(students_db)

# Import the loans table
loans_db <- dbGetQuery(con_sqlite, "SELECT * FROM loans")
head(loans_db)

# Query with a filter — students with math_score >= 15
high_scores <- dbGetQuery(con_sqlite,
  "SELECT studentID, gender, math_score, it_score
   FROM students
   WHERE math_score >= 15")
print(high_scores)

dbDisconnect(con_sqlite)
```

### 4.2 PostgreSQL (requires a running PostgreSQL server)

```{r postgres, eval=FALSE}
install.packages("RPostgres")
library(RPostgres)

con_pg <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "your_database_name",
  host     = "localhost",
  user     = "your_username",
  password = "your_password",
  port     = 5432
)

students_pg <- dbGetQuery(con_pg, "SELECT * FROM students LIMIT 10")
head(students_pg)

dbDisconnect(con_pg)
```

### 4.3 MySQL (requires a running MySQL server)

```{r mysql, eval=FALSE}
install.packages("RMySQL")
library(RMySQL)

con_mysql <- dbConnect(
  RMySQL::MySQL(),
  dbname   = "your_database_name",
  host     = "localhost",
  user     = "your_username",
  password = "your_password",
  port     = 3306
)

students_mysql <- dbGetQuery(con_mysql, "SELECT * FROM students LIMIT 10")
head(students_mysql)

dbDisconnect(con_mysql)
```

---

## Summary

| Source                  | Package    | Function           |
|-------------------------|------------|--------------------|
| Text file (.txt)        | base R     | `read.table()`     |
| CSV file (.csv)         | base R     | `read.csv()`       |
| Excel file (.xlsx)      | `readxl`   | `read_excel()`     |
| SPSS file (.sav)        | `haven`    | `read_sav()`       |
| Stata file (.dta)       | `haven`    | `read_dta()`       |
| SAS file (.sas7bdat)    | `haven`    | `read_sas()`       |
| SQLite database         | `DBI` + `RSQLite`  | `dbGetQuery()` |
| PostgreSQL database     | `DBI` + `RPostgres`| `dbGetQuery()` |
| MySQL database          | `DBI` + `RMySQL`   | `dbGetQuery()` |
